---
title: "The shakerr Package"
author: "Steven Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The shakerr Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message= FALSE}

require(dplyr)
require(shakerr)
require(ggplot2)

```


## Introduction

This package offers functionality to format earthquake data sets downloaded from
the U.S. National Oceanographic and Atmospheric Administration (NOAA).

Once the data set is formatted the package offers two ways to visualize the data.
The first way is by using a timeline graph where the magnitude and year of the
earthquake is graphed. The second is by mapping the epicenters of a selection 
of earthquakes directly on a map.

The plots will not only show the location but the magnitude as well. 
The line graphs will have two sub-types, one that labels the top earthquakes
in the timeline and one that does not. Likewise the epicenter graphs will
have two subtypes, one that gives basic information about the earthquake and
one with a little more detail of the magnitude and number of deaths.

This vignette will be split into two parts, the first will describe the
cleaning and formatting of the data and the second the visualization of the 
data.

## Cleaning and Formatting the Data

There are three types of sample data available to demonstrate the functionality
in this package. An example of one is the `eq_test_data` data set. This data 
set contains the magnitude and other attributes of 100 earthquakes from around 
the world since recorded time of the NOAA database. The information for these 
data sets may be read by running a help call `?eq_test_data` on the data set.

The following demonstrates how the data set is formatted using the package 
functionality.

`eq_clean_data`: creates a Date classed field with the appropriate date format,
and converts the longitude and latitude fields to a numeric class.
`eq_location_clean`: reformats the location name of the earthquake.

```{r clean and formatted}
#Data before formatting
head(select(eq_test_data, HOUR, MINUTE, SECOND, LONGITUDE, LATITUDE,
                   LOCATION_NAME))
#Formatting the data
frm_data <- eq_clean_data(eq_test_data) %>% eq_location_clean()

#Formatted data
head(select(frm_data, DATE, LONGITUDE, LATITUDE, LOCATION_NAME))


```

Note the first data set shows the unformatted data and the second the formatted
data.

## Data Visualization

The formatted data can be visualized into two ways, the first as a timeline and
the second as points of occurence on a map.

The following functions show how a timeline may be graphed, with and 
without labels. The function `quaketimeline_plot` utilises the `GeomTimeline`
and `StatTimeline` Geom and Stat respectively. Labelling the timelines requires
`quaketimelinelabel_plot` which uses the `GeomTimelinelable` and 
`StatTimelinelabel` Geom and Stat respectively. Note that with labels,
the extra parameter indicates the top number of earthquakes to label by
magnitude.

```{r timeline graphs without labels}

quaketimeline_plot(eq_clean_testdata[1:12,], "1800-01-01","2014-01-01")

```

Instead of using the function, the underlying `geom_timeline` function may be
used. Note that a minimum and maximum date need to be entered.

```{r geom_timeline function}

ggplot(eq_china_cleandata, aes( x = DATE, y = as.factor(COUNTRY))) + 
  geom_timeline( x_min = "1800-01-01", x_max = "2015-01-01")

```



```{r timeline with labels}

quaketimelinelabel_plot(eq_china_cleandata, "1800-01-01","2015-01-01", 3)

```

Again, going directly to the underlying function `geom_timelinelabel` is 
possible if a little more flexibility is required.

```{r}

ggplot(eq_china_cleandata, aes( x = DATE, y = as.factor(COUNTRY))) +
  geom_timeline( x_min = "1800-01-01", x_max = "2015-01-01") +
  geom_timelinelabel(ggplot2::aes_(magnitude = quote(EQ_PRIMARY),
                                   label = quote(LOCATION_NAME),
                                   col = NULL),
                     x_min = "1800-01-01",
                     x_max = "2015-01-01",
                     n_max = 3)

```


The second kind of visualization plots the epicenters of the selected 
earthquakes on a map. The epicenters diameter will represent the magnitude
of the earthquake, with the each data point showing further information when
you click on it. The first graph will present the date of the earthquake
when you click on it. The second will offer more detail, with location and 
magnitude.

```{r epicenter mapping}

eq_map(eq_china_cleandata, "DATE")

```


```{r epicenter mapping with info}

  eq_clean_data(eq_test_data) %>% 
  dplyr::filter(COUNTRY == "TURKEY" & lubridate::year(DATE) >= 1500) %>% 
  dplyr::mutate(popup_text = eq_create_label(.)) %>% 
  eq_map(annot_col = "popup_text")

```

